\textbf{Introducing the Choice-Confidence (CHOCO) Model for Bimodal Data from Subjective Ratings: Application to the Effect of Attractiveness on Reality Beliefs about AI-Generated Faces}

### 2025-06-21 paper/manuscript.tex
update based on comments

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, arc=.35mm, colback=white, rightrule=.15mm, leftrule=.75mm, left=2mm, breakable, colframe=quarto-callout-note-color-frame, opacityback=0, toprule=.15mm]

@@

\end{tcolorbox}  Author roles were classified using the Contributor Role
Taxonomy (CRediT; https://credit.niso.org/) as follows:  Dominique
Makowski:   Conceptualization, Data curation, Formal Analysis, Funding
acquisition, Investigation, Methodology, Project
administration, Resources, Software, Supervision, Validation, Visualization, Writing
-- original draft; Ana Neves:   Data curation, Writing -- original
draft, Writing -- review \& editing; Andy Field:   Writing -- original
draft, Writing -- review \& editing}
\par{Correspondence concerning this article should be addressed
to Dominique
Makowski, Email: \href{mailto:D.Makowski@sussex.ac.uk}{D.Makowski@sussex.ac.uk}}
}

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-blanca2018current}{Blanca et al., 2018};
\citeproc{ref-cumming2014new}{Cumming, 2014};
\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data\footnote{More specifically, that the outcome
  is distributed according to a Normal distribution which parameters are
  expressed as a linear function of the predictors.} - as this
assumption often does not hold true for many types of psychological
outcomes. For instance, reaction times typically exhibit skewed
distributions, choices can be represented as binary variables, and count
data consists of strictly positive integers. Applying models that
presume normality and model the ``mean'' of the outcome variable can
lead to misinterpretation and potentially misleading conclusions when
applied indiscriminately. It is thus important that psychologists use
models that can best describe (or generate) the data they collect, to
fully exploit them and bring more nuance and accuracy to their
conclusions.

Among the most commonly collected data in psychology are responses on
subjective scales, such as Likert-type items or visual analog scales,
which exhibit some fundamental properties: these responses are bounded
(and can be rescaled to a 0-1 range) and frequently display clustering
at the extremes. Traditional linear models being ill-suited for such
data, researchers have turned to using Beta distributions to model this
data (instead of Gaussian), suited for continuous data within the (0,1)
interval (i.e., excluding extreme responses). To address the frequent
occurrence of exact zeros and ones (i.e., extreme values), zero-one
inflated beta (ZOIB) models have been developed
(\citeproc{ref-ospina2012general}{Ospina \& Ferrari, 2012}) to
accommodate the excess of boundary values by incorporating additional
components that model the probabilities of responses at 0 and 1 as a
separate, independent process.

@@

The Beta-Gate model is a reparametrized Ordered Beta model
(\citeproc{ref-kubinec2023ordered}{Kubinec, 2023})\footnote{In the
  Ordered Beta model, the cutpoints on the log-scale are directly used
  as parameters, instead of being derived from \emph{pex} and
  \emph{bex}.} available in the \emph{cogmod} package in R
(https://github.com/DominiqueMakowski/cogmod), in which participants'
answers on bounded scales are conceptualized as latent responses that
can fall past a pair of probabilistic ``gates'' (or cutpoints) that
control whether the response is recorded as an extreme (0 or 1) or as a
nuanced, continuous value in between (Figure~\ref{fig-one}). These
distance of these gates from the edges of the scale varies based on two
interpretable parameters: \emph{pex} (the propensity by which people are
likely to answer extreme values), and \emph{bex}, a bias toward the
upper extreme (1) versus the lower (0). A person's internal response
that lies close to the edge might be ``caught'' by a gate and recorded
as an extreme, while others pass through to express a continuous
response (Beta-distributed with \(\mu\) (\emph{mu}) and \(\phi\)
(\emph{phi}) as its mean and precision parameters). The Beta-Gate model
is based on the idea that extreme values can emerge not just from a
fundamentally different underlying processes - as assumed in ZOIB models
- but from a common process governed by thresholds of decisiveness and
confidence.

@@

\caption{\label{fig-one}The Beta-Gate Distribution is a reparametrized
ordered Beta model (Kubinec, 2023) that is governed by 4 parameters.
`Mu' and `phi' correspond to the mean and precision of the continuous
part of the distribution (between 0 and 1), and `pex' (propensity of
extremes) and `bex' (balance of extremes) indirectly control the
proportion of zeros and ones by specifying the location of the
``gates'', past which the latent response process is likely to generate
extreme values. Specifically, `pex' defines the total distance of both
gates from the extremes (in yellow), and `bex' determines the proportion
of the right gate distance relative to the left. In this example, the
total distance from the extremes is `pex' = 0.1, with 40\% (`bex' = 0.4)
of that distance being on the right (and 60\% on the left). The left
gate is thus located at 0.6, and the right at 1-0.04 = 0.96.}

@@

\begin{itemize}
\tightlist
\item
  \(P(x = 0) = \text{logistic}\left( \text{logit}(pex \cdot (1 - bex)) - \text{logit}(\mu) \right)\)
\item
  \(P(x = 1) = 1 - \text{logistic}\left( \text{logit}(1 - pex \cdot bex) - \text{logit}(\mu) \right)\)
\item
  \(P(x \in (0,1)) = 1 - P(x = 0) - P(x = 1)\)
\end{itemize}

@@

\caption{\label{fig-two}The CHOCO Model uses a mixture of Beta-Gate
distributions to model separately the right and left sides of the scale
(e.g., a rating of whether a statement was `Truth' vs.~`Lie'), as well
as their relative proportion. In this example, the participants are more
likely overall (p = 60\%) to select the right side of the scale (`Lie')
than the left (`Truth'). They are also more confident in their choice
(confright = 0.6 vs.~confleft = 0.3). Extreme values (zeros and ones)
are governed by the same mechanism as for Beta-Gate models.}

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig2.png}

@@

\textbf{Study 1} aims to evaluate the CHOCO model's ability to better
capture subjective scale responses that (potentially) reflect an
underlying discrete choice, in comparison to existing models such as the
ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO
provides improved model fit, 2) yields deeper insights into
population-level effects than traditional approaches (gender differences
in reality beliefs), and 3) allows for the reliable estimation of
interpretable individual-level parameters through random effects.
\textbf{Study 2} will apply this model to more subtle effects, such as
the effect perceived facial attractiveness on reality judgments, and
test the ability to fit alternative data structures (scales with ordinal
response options and mid-points). To this end, we analyze data from two
separate studies in which participants judged whether a face image was
AI-generated (``fake'') or a real photograph.

@@

In today's post-truth era, the proliferation of advanced AI technologies
has made it increasingly challenging to distinguish between authentic
and synthetic media, bearing significant implications for information
integrity and public trust
(\citeproc{ref-lewandowsky2017beyond}{Lewandowsky et al., 2017}). As
traditional cues become less reliable (e.g., visual glitches and
artefacts in generated images; formulaic generated text, etc.), people
increasingly depend on contextual information and cognitive heuristics
to assess authenticity, a process referred to as ``simulation
monitoring'' (\citeproc{ref-makowski2019phenomenal}{Makowski, Sperduti,
et al., 2019}).

@@

Using the open-access data from Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}), we included all heterosexual and
bisexual (as these two groups did not seem to differ based on
preliminary analyses and were thus grouped to maximize power) male and
female participants, for a final sample of 141 participants (Mean age =
28.4, SD = 9.0, range: {[}19, 66{]}; Sex: 47.5\% females). For each
participant, we included only stimuli of the opposite gender (i.e., all
89 female faces for men and 20 male faces for women).

@@

In the first phase, participants viewed 109 neutral-expression
photographs of faces (random order, display time of 3 s) from the
American Multiracial Face Database (AMFD,
\citeproc{ref-chen2021broadening}{Chen et al., 2021}). After each image,
participants rated the face on trustworthiness, familiarity,
attractiveness, and beauty using visual analog scales. In the second
phase, participants were informed that ``about half of the previously
seen images were AI-generated''. The same faces were presented again in
a new random order (same display time), followed by ratings of
``reality'' (whether they believed the image was fake - left anchor - or
real - right anchor).

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a
Beta-Gate model, and the CHOCO model. For all models and each parameter,
the full formula was entered:
\(Real\sim Sex + (1 | Participant) + (1 | Item)\) (with Sex as the main
predictor and participants and items entered as random intercepts). The
models were run using \emph{brms}
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) R package, and analyzed
using the \emph{easystats} collection of packages
(\citeproc{ref-ludecke2020extracting}{Lüdecke et al., 2020};
\citeproc{ref-makowski2025modelbased}{Makowski, Ben-Shachar, et al.,
2025}; \citeproc{ref-patil2022datawizard}{Patil et al., 2022}). To
maximize the comparability across models We used the default priors
(uniform) for all models, and we ran 16 chains of 1400 iterations each
on the University of Sussex High-Performance Computing (HPC) cluster.

@@

For the population-level effects, we will consider significant and
report (using the median of the posterior distribution) effects for
which the 95\% Credible Interval (CI) does not include zero (and when
the probability of direction \emph{pd} is \textgreater{}
\textasciitilde97\%, \citeproc{ref-makowski2019indices}{Makowski,
Ben-Shachar, et al., 2019}). For the individual-level parameters (i.e.,
the random intercepts of each parameter for each participant and each
item), we will first analyze their reliability using the
Variance-Over-Uncertainty Ratio index (\emph{D-vour}). This index,
implemented in the \emph{performance} package
(\citeproc{ref-ludecke2021performance}{Lüdecke et al., 2021}), is
inspired by recent work on mixed models reliability
(\citeproc{ref-rouder2024hierarchical}{Rouder \& Mehrvarz, 2024};
\citeproc{ref-williams2021beneath}{Williams et al., 2021}), and
corresponds to the normalized ratio of observed variability to
uncertainty in random effect estimates, defined as:

@@

Where \(\sigma_{between}^2\) is the between-group variability (computed
as the SD of the random effect point-estimates) and
\(\mu_{\text{within}}^2\) is the mean squared uncertainty (SE) in random
effect estimates (i.e., the average uncertainty). We use as
\emph{D-vour} = 0.666 as the threshold for moderately reliable random
effect estimates, which corresponds to a 2:1 ratio of between-group
variance to uncertainty.

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig3.png}

@@

The ZOIB model suggested that women had higher mean scores of reality
beliefs
(\(\mu_{~\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], ~pd = 98.77\%\)),
less extreme values
(\(zoi_{~\text{Female}} = -1.24,~95\%~ CI~[-2.46, -0.07], ~pd = 98.25\%\))
but more ones relative to zeros
(\(coi_{~\text{Female}} = 1.63,~95\%~ CI~[0.70, 2.56], ~pd = 99.97\%\)).

The Beta-Gate model similarly suggested that women had higher mean
scores of reality beliefs
(\(\mu_{~\text{Female}} = 0.21,~95\%~ CI~[0.02, 0.40], ~pd = 98.52\%\)),
less extreme values
(\(pex_{~\text{Female}} = -1.34,~95\%~ CI~[-2.56, -0.17], ~pd = 98.75\%\)),
and a greater tendency to answer one relative to zero
(\(bex_{~\text{Female}} = 1.13,~95\%~ CI~[0.34, 1.91], ~pd = 99.76\%\)).

The CHOCO model shows that women had a higher probability \emph{p} of
judging faces as real
(\(P_{~\text{Female}} = 0.45,~95\%~ CI~[0.05, 0.86], ~pd = 98.54\%\)),
but are not more confident when doing so
(\(confright_{~\text{Female}} = -0.13,~95\%~ CI~[-0.47, 0.21], ~pd = 77.73\%\)).
However, they were less confident when answering that an image was
AI-generated
(\(confleft_{~\text{Female}} = -0.53,~95\%~ CI~[-0.89, -0.17], ~pd = 99.84\%\)).
There were also less likely to produce extreme answers
(\(pex_{~\text{Female}} = -1.19,~95\%~ CI~[-2.40, 0.01], ~pd = 97.64\%\)),
but no strong evidence supporting a directional bias was observed
(\(bex_{~\text{Female}} = 0.48,~95\%~ CI~[-0.12, 1.09], ~pd = 94.32\%\)).

@@

{\caption{{Top: the participant and item-level estimates for the CHOCO
parameters, under the distribution of their point-estimates. Reliable
effects are characterised by a higher between-group variability relative
(the dispersion of the distribution of point-estimates) to its
within-group variability (the average uncertainty of individual
estimates represented by the error bar). For example, inter-individual
variability in the parameters reflecting the confidence in left and
right choices is reliably capture, as opposed to the inter-individual
variability in the `bex' parameter. Bottom: correlation matrix between
participant-level estimates from different models and empirical indices
(e.g., the raw proportion of responses on the right per participant).
The CHOCO parameters are easily interpretable, strongly correlating with
their empirical counterparts.}{\label{fig-four}}}}

\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./figures/fig4.png}

@@

The explosion of accessibility of state-of-the-art AI tools has made it
effortless to generate realistic images, including human faces that are
often indistinguishable from real ones
(\citeproc{ref-bozkir2024can}{Bozkir et al., 2024};
\citeproc{ref-miller2023ai}{Miller et al., 2023};
\citeproc{ref-nightingale2022ai}{Nightingale \& Farid, 2022}). These
synthetic visuals are flooding the cyberspace across various domains,
such as art, advertising and entertainment, to education and
information. This technological advancement carries an important
potential for misuse, such as in disinformation campaigns, scams (e.g.,
AI-bots, identity theft), and abuse. The democratization of such
technology raises pressing concerns about the value of authenticity and
the potential erosion of media trust in our increasingly
\emph{post-truth} society.

In this evolving landscape, understanding the cognitive mechanisms that
underpin our judgments of reality becomes paramount. Despite the
increasing prevalence of digitally altered or AI-generated content,
humans still rely on certain heuristics to assess authenticity. One such
heuristic might be facial attractiveness. Attractiveness appraisals are
known to be automatic and unconscious (\citeproc{ref-hou2023neural}{Hou
et al., 2023}; \citeproc{ref-hung2016unconscious}{Hung et al., 2016};
\citeproc{ref-luo2019robust}{Luo et al., 2019}), and carry strong
real-life consequences, as demonstrated by the large body of literature
on the ``beauty premium''\footnote{As an eloquent example, Monk Jr et
  al. (\citeproc{ref-monk2021beholding}{2021}) reported in a large
  representative US sample that the magnitude of earnings disparities
  among white women along the perceived attractiveness continuum exceeds
  in magnitude the canonical black-white race gap.}
(\citeproc{ref-gulati2024beautiful}{Gulati et al., 2024};
\citeproc{ref-kukkonen2024beauty}{Kukkonen et al., 2024};
\citeproc{ref-little2021facial}{Little, 2021};
\citeproc{ref-pandey2021face}{Pandey \& Zayas, 2021}).

Although its role as a potential modulator of reality beliefs remains
under-explored, Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}) found significant associations
between participants' realness ratings and facial attractiveness, with a
potential sexual dymorphism: male participants judged attractive faces
as being likely more real, whereas evidence suggested a milder and
quadratic (U-shaped) relationship between attractiveness and reality
beliefs for females. These findings offer a complementary perspective to
those of Miller et al. (\citeproc{ref-miller2023ai}{2023}), who reported
that participants used attractiveness as a distinguishing cue between
real and AI-generated faces. These results highlight a possible
bidirectional and context-dependent influence of attractiveness on
reality judgments, of which the exact shape needs further investigation.

@@

The first sample includes the same participants as study 1 (N = 141, see
above). For the second sample (data and preprocessing information
available at https://github.com/RealityBending/FakeFace2), 248
participants were initially recruited through academic platforms (SONA
and SurveySwap). We removed 15 participants with data suggesting
low-effort responding as well as those (N=16) that did not believe in
the experimental manipulation and were fully confident that all images
were real or fake. Similarly to the first sample, we included all hetero
and bisexual participants and stimuli from the opposite gender,
resulting in a final sample included 189 participants (Mean age = 28.4,
SD = 14.0, range: {[}18, 69{]}; Sex: 76.2\% females). The study was
approved by the University of Sussex Ethics Board (ER/ST633/1).

@@

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig5.png}

@@

\subsubsection{Beauty}\label{beauty}

The same analysis was run for Beauty, which indicated the following
differences: \textbf{ANA: can you check and write down the main
differences?}

@@

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Andy Field}{true}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, opacityback=0, left=2mm, leftrule=.75mm, colback=white, breakable, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, toprule=.15mm, arc=.35mm, rightrule=.15mm]

@@

\end{tcolorbox}  Author roles were classified using the Contributor Role
Taxonomy (CRediT; https://credit.niso.org/) as follows:  Dominique
Makowski:   Conceptualization, Data curation, Formal Analysis, Funding
acquisition, Investigation, Methodology, Project
administration, Resources, Software, Supervision, Validation, Visualization, Writing
-- original draft; Ana Neves:   Data curation, Writing -- original
draft, Writing -- review \& editing; Andy Field:   Writing -- review \&
editing}
\par{Correspondence concerning this article should be addressed
to Dominique
Makowski, Email: \href{mailto:D.Makowski@sussex.ac.uk}{D.Makowski@sussex.ac.uk}}
}

@@

Despite significant advancements in psychological science following the
replication crisis (\citeproc{ref-open2015estimating}{Collaboration,
2015}), its progress is still hindered by its sub-optimal (or
inappropriate) usage of statistical tools
(\citeproc{ref-blanca2018current}{Blanca et al., 2018};
\citeproc{ref-cumming2014new}{Cumming, 2014};
\citeproc{ref-makowski2023we}{Makowski \& Waggoner, 2023}). A prevalent
issue is the continued reliance on linear models that assume normally
distributed (Gaussian) data\footnote{More specifically, that the outcome
  is distributed according to a Normal distribution which parameters are
  expressed as a linear function of the predictors.} - as this
assumption often does not hold true for many types of psychological
outcomes. For instance, reaction times typically exhibit skewed
distributions, choices can be represented as binary variables, and count
data consist of strictly positive integers. Applying models that presume
normality and model the ``mean'' of the outcome variable can lead to
misinterpretation and potentially misleading conclusions when applied
indiscriminately. It is thus important that psychologists use models
that can best describe (or generate) the data they collect, to exploit
them fully and bring more nuance and accuracy to their conclusions.

Among the most commonly collected data in psychology are responses on
subjective scales, such as Likert-type items or visual analog scales,
which exhibit some fundamental properties: these responses are bounded
(and can be rescaled to a 0-1 range) and frequently display clustering
at the extremes. Traditional linear models being ill-suited for such
data, researchers have turned to using Beta distributions to model these
data (instead of Gaussian), suited for continuous data within the (0,1)
interval (i.e., excluding extreme responses). To address the frequent
occurrence of exact zeros and ones (i.e., extreme values), zero-one
inflated beta (ZOIB) models have been developed
(\citeproc{ref-ospina2012general}{Ospina \& Ferrari, 2012}) to
accommodate the excess of boundary values by incorporating additional
components that model the probabilities of responses at 0 and 1 as a
separate, independent process.

@@

The Beta-Gate model is a reparametrized Ordered Beta model
(\citeproc{ref-kubinec2023ordered}{Kubinec, 2023})\footnote{In the
  Ordered Beta model, the cutpoints on the log-scale are directly used
  as parameters, instead of being derived from \emph{pex} and
  \emph{bex}.} available in the \emph{cogmod} package in R
(https://github.com/DominiqueMakowski/cogmod), in which participants'
answers on bounded scales are conceptualized as latent responses that
can fall past a pair of probabilistic ``gates'' (or cutpoints) that
control whether the response is recorded as an extreme (0 or 1) or as a
nuanced, continuous value in between (Figure~\ref{fig-one}). The
distance of these gates from the edges of the scale varies based on two
interpretable parameters: \emph{pex} (the propensity by which people are
likely to answer extreme values), and \emph{bex}, a bias toward the
upper extreme (1) versus the lower (0). A person's internal response
that lies close to the edge might be ``caught'' by a gate and recorded
as an extreme, while others pass through to express a continuous
response (Beta-distributed with \(\mu\) (\emph{mu}) and \(\phi\)
(\emph{phi}) as its mean and precision parameters). The Beta-Gate model
is based on the idea that extreme values can emerge not just from a
fundamentally different underlying processes - as assumed in ZOIB models
- but from a common process governed by thresholds of decisiveness and
confidence.

@@

\caption{\label{fig-one}The Beta-Gate Distribution is a reparametrized
ordered Beta model (Kubinec, 2023) that is governed by 4 parameters.
`Mu' and `phi' correspond to the mean and precision of the continuous
part of the distribution (between 0 and 1), and `pex' (propensity of
extremes) and `bex' (balance of extremes) indirectly control the
proportion of zeros and ones by specifying the location of the
``gates'', past which the latent response process is likely to generate
extreme values. Specifically, `pex' defines the total distance of both
gates from the extremes (i.e., the sum length of the two orange arrows),
and `bex' determines the proportion of the right gate distance relative
to the left (i.e., the length of the right arrow relative to the left
one). In this example, the total distance from the extremes is `pex' =
0.1, with 40\% (`bex' = 0.4) of that distance being on the right (and
60\% on the left). The left gate is thus located at 0.6, and the right
at 1-0.04 = 0.96.}

@@

\[
\begin{cases}
\mathrm{logistic}\left( \mathrm{logit}(pex \cdot (1 - bex)) - \mathrm{logit}(\mu) \right), & x = 0, \\[6pt]
1 - \mathrm{logistic}\left( \mathrm{logit}(1 - pex \cdot bex) - \mathrm{logit}(\mu) \right), & x = 1, \\[6pt]
1 - P(x = 0) - P(x = 1), & 0 < x < 1. \\
\end{cases}
\]

@@

{\caption{{Top: The CHOCO Model uses a mixture of Beta-Gate
distributions to model separately the right and left sides of the scale
(e.g., a rating of whether a statement was `Truth' vs.~`Lie'), as well
as their relative proportion. In this example, the participants are more
likely overall (p = 60\%) to select the right side of the scale (`Lie')
than the left (`Truth'). They are also more confident in their choice
(confright = 0.6 vs.~confleft = 0.3). Extreme values (zeros and ones)
are governed by the same mechanism as for Beta-Gate models. Below: An
illustrative dataset in which participants, in condition A (blue),
respond as often on the left and right side of the scale, with low
confidence (the confidence is concentrated towards the centre). In
condition B (orange), they answer more often on the left (but also with
low confidence), but their responses on the right are more extreme. As
shown in the top table, a typical linear model could lead the imprudent
researcher to misleadingly conclude that ``there is no effect of
Condition'' (as a linear regression models changes in the mean, which is
actually equal for the two conditions). The CHOCO model correctly
reveals that condition B is associated with a lower probability but
higher confidence of answers on the right side of the scale. Note that
the CHOCO model is currently only available through Bayesian modeling,
hence we report for consistency pseudo t- and p-values in the table,
which correspond respectively to the coefficients divided by their SE,
and the values obtained via the pd-to-p conversion
(\citeproc{ref-makowski2019indices}{Makowski, Ben-Shachar, et al.,
2019}).}{\label{fig-two}}}}

\begin{center}
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{./figures/fig2final.png}
\end{center}

@@

\textbf{Study 1} aims to evaluate the CHOCO model's ability to better
capture subjective scale responses that (potentially) reflect an
underlying discrete choice, in comparison to existing models such as the
ZOIB and Beta-Gate. Specifically, we will assess whether 1) CHOCO
provides improved model fit, 2) yields deeper insights into
population-level effects than traditional approaches (sex differences in
reality beliefs), and 3) allows for the reliable estimation of
interpretable individual-level parameters through random effects.
\textbf{Study 2} will apply this model to more subtle effects, such as
the effect of perceived facial attractiveness on reality judgments, and
test the ability to fit alternative data structures (scales with ordinal
response options and mid-points). To this end, we analyze data from two
separate studies in which participants judged whether a face image was
AI-generated (``fake'') or a real photograph.

@@

The question of ``what is real'' has become increasingly relevant in the
context of recent technological developments, particularly with the rise
of artificial intelligence (AI), capable of mass producing highly
realistic and relevant content, such as images, videos or text. These
synthetic productions are flooding the cyberspace across various
domains, such as art, advertising and entertainment, to education and
information. The democratization of such technology carries an important
potential for misuse, such as in disinformation campaigns and scams
(e.g., AI-bots, identity theft), and abuse (e.g., deepfakes,
\citeproc{ref-viola2023designed}{Viola \& Voto, 2023}), and raises
pressing concerns about the value of authenticity and the potential
erosion of media trust in our increasingly \emph{post-truth} society
(\citeproc{ref-lewandowsky2017beyond}{Lewandowsky et al., 2017}).
Critically, as technology improves, traditional cues for detecting what
is fake become less reliable (e.g., visual glitches and artefacts in
generated images; formulaic generated text, etc.), leading people to
increasingly depend on contextual information and cognitive heuristics
to assess authenticity, a process referred to as ``simulation
monitoring'' (\citeproc{ref-makowski2019phenomenal}{Makowski, Sperduti,
et al., 2019}).

@@

Using the open-access data from Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}), we included all heterosexual and
bisexual (as these two groups did not seem to differ based on
preliminary analyses and were thus grouped to maximize power) male and
female participants, for a final sample of 141 participants (Mean age =
28.4, SD = 9.0, range: {[}19, 66{]}; Sex: 47.5\% females). For each
participant, we included only stimuli of the opposite sex (i.e., all 89
female faces for males and 20 male faces for females).

@@

In the first phase, participants viewed 109 neutral-expression
photographs of faces (random order, display time of 3 s) from the
American Multiracial Face Database (AMFD,
\citeproc{ref-chen2021broadening}{Chen et al., 2021}). After each image,
participants rated the face on trustworthiness, familiarity,
attractiveness, and beauty using visual analog scales. In the second
phase, participants were informed that ``about half of the previously
seen images were AI-generated'' (in actuality, all faces were
photographs). The same faces were presented again in a new random order
(same display time), followed by ratings of ``reality'' (whether they
believed the image was fake - left anchor - or real - right anchor).

@@

We fitted 3 models to predict the reality ratings: a ZOIB model, a
Beta-Gate model, and the CHOCO model. For all models and each parameter,
the full formula was entered:
\(Real\sim Sex + (1 | Participant) + (1 | Item)\) (with Sex as the main
predictor and participants and items entered as random intercepts). The
models were run using \emph{brms}
(\citeproc{ref-burkner2017brms}{Bürkner, 2017}) R package, and analyzed
using the \emph{easystats} collection of packages
(\citeproc{ref-ludecke2020extracting}{Lüdecke et al., 2020};
\citeproc{ref-makowski2025modelbased}{Makowski, Ben-Shachar, et al.,
2025}; \citeproc{ref-patil2022datawizard}{Patil et al., 2022}). To
maximize comparability across models, we used the default priors
(uniform) for all models, and we ran 16 chains of 1400 iterations each
on the University of Sussex High-Performance Computing (HPC) cluster.

@@

For the population-level effects, we will consider ``significant''
(noteworthy) and report (using the median of the posterior distribution)
effects for which the 95\% Credible Interval (CI) does not include zero
(and when the probability of direction \emph{pd} is \textgreater{}
\textasciitilde97\%, \citeproc{ref-makowski2019indices}{Makowski,
Ben-Shachar, et al., 2019}). For the individual-level parameters (i.e.,
the random intercepts of each parameter for each participant and each
item), we will first analyze their reliability using the
Variance-Over-Uncertainty Ratio index (\emph{D-vour}). This index,
implemented in the \emph{performance} package
(\citeproc{ref-ludecke2021performance}{Lüdecke et al., 2021}), is
inspired by recent work on mixed models reliability
(\citeproc{ref-rouder2024hierarchical}{Rouder \& Mehrvarz, 2024};
\citeproc{ref-williams2021beneath}{Williams et al., 2021}), and
corresponds to the normalized ratio of observed variability to
uncertainty in random effect estimates, defined as:

@@

Where \(\sigma_{between}^2\) is the between-group variability (computed
as the SD of the random effect point-estimates) and
\(\mu_{\text{within}}^2\) is the mean squared uncertainty (SE) in random
effect estimates (i.e., the average uncertainty). We use as
\emph{D-vour} = 0.666 as the threshold for moderately reliable random
effect estimates, which corresponds to a 2:1 ratio of between-group
variance to uncertainty. Reliable in this context means that the
variability between groups (e.g., participants) is greater than the
uncertainty in the estimates, suggesting that the parameter can be used
to assess inter-individual differences.

@@

\begin{center}
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig3.png}
\end{center}

@@

The ZOIB model suggested that females had higher mean scores of reality
beliefs
(\(\mu_{~\text{Female}} = 0.20,~95\%~ CI~[0.03, 0.37], ~pd = 98.77\%\)),
less extreme values
(\(zoi_{~\text{Female}} = -1.24,~95\%~ CI~[-2.46, -0.07], ~pd = 98.25\%\))
but more ones relative to zeros
(\(coi_{~\text{Female}} = 1.63,~95\%~ CI~[0.70, 2.56], ~pd = 99.97\%\)).

The Beta-Gate model similarly suggested that females had higher mean
scores of reality beliefs
(\(\mu_{~\text{Female}} = 0.21,~95\%~ CI~[0.02, 0.40], ~pd = 98.52\%\)),
less extreme values
(\(pex_{~\text{Female}} = -1.34,~95\%~ CI~[-2.56, -0.17], ~pd = 98.75\%\)),
and a greater tendency to answer one relative to zero
(\(bex_{~\text{Female}} = 1.13,~95\%~ CI~[0.34, 1.91], ~pd = 99.76\%\)).

The CHOCO model shows that females had a higher probability \emph{p} of
judging faces as real
(\(P_{~\text{Female}} = 0.45,~95\%~ CI~[0.05, 0.86], ~pd = 98.54\%\)),
but are not more confident when doing so
(\(confright_{~\text{Female}} = -0.13,~95\%~ CI~[-0.47, 0.21], ~pd = 77.73\%\)).
However, they were less confident when answering that an image was
AI-generated
(\(confleft_{~\text{Female}} = -0.53,~95\%~ CI~[-0.89, -0.17], ~pd = 99.84\%\)).
There were also less likely to produce extreme answers
(\(pex_{~\text{Female}} = -1.19,~95\%~ CI~[-2.40, 0.01], ~pd = 97.64\%\)),
but no strong evidence supporting a directional bias was observed
(\(bex_{~\text{Female}} = 0.48,~95\%~ CI~[-0.12, 1.09], ~pd = 94.32\%\)).

@@

{\caption{{Top: the participant and item-level estimates for the CHOCO
parameters, under the distribution of their point-estimates. Reliable
effects are characterised by a higher between-group variability relative
(the dispersion of the distribution of point-estimates) to its
within-group variability (the average uncertainty of individual
estimates represented by the error bar). For example, inter-individual
variability in the parameters reflecting the confidence in left and
right choices is reliably captured, as opposed to the inter-individual
variability in the `bex' parameter. Bottom: correlation matrix between
participant-level estimates from different models and empirical indices
(e.g., the raw proportion of responses on the right per participant).
The CHOCO parameters are easily interpretable, strongly correlating with
their empirical counterparts.}{\label{fig-four}}}}

\begin{center}
\includegraphics[width=0.8\linewidth,height=\textheight,keepaspectratio]{./figures/fig4.png}
\end{center}

@@

The explosion of accessibility of state-of-the-art AI tools has made it
effortless to generate realistic images, including human faces that are
often indistinguishable from real ones
(\citeproc{ref-bozkir2024can}{Bozkir et al., 2024};
\citeproc{ref-miller2023ai}{Miller et al., 2023};
\citeproc{ref-nightingale2022ai}{Nightingale \& Farid, 2022}). In this
evolving landscape, understanding the cognitive mechanisms that underpin
our judgments of reality becomes paramount. Despite the increasing
prevalence of digitally altered or AI-generated content, humans still
rely on certain heuristics to assess authenticity. One such heuristic
might be facial attractiveness. Attractiveness appraisals are known to
be automatic and unconscious (\citeproc{ref-hou2023neural}{Hou et al.,
2023}; \citeproc{ref-hung2016unconscious}{Hung et al., 2016};
\citeproc{ref-luo2019robust}{Luo et al., 2019}), and carry strong
real-life consequences, as demonstrated by the large body of literature
on the ``beauty premium''\footnote{As an eloquent example, Monk Jr et
  al. (\citeproc{ref-monk2021beholding}{2021}) reported in a large
  representative US sample that the magnitude of earnings disparities
  among white women along the perceived attractiveness continuum exceeds
  in magnitude the canonical black-white race gap.}
(\citeproc{ref-gulati2024beautiful}{Gulati et al., 2024};
\citeproc{ref-kukkonen2024beauty}{Kukkonen et al., 2024};
\citeproc{ref-little2021facial}{Little, 2021};
\citeproc{ref-pandey2021face}{Pandey \& Zayas, 2021}).

Although the role of attractiveness as a potential modulator of reality
beliefs remains under-explored, Makowski, Te, et al.
(\citeproc{ref-makowski2025too}{2025}) found significant associations
between participants' realness ratings and facial attractiveness, with a
potential sexual dymorphism: male participants judged attractive faces
as being likely more real, whereas evidence suggested a milder and
quadratic (U-shaped) relationship between attractiveness and reality
beliefs for females. These findings offer a complementary perspective to
those of Miller et al. (\citeproc{ref-miller2023ai}{2023}), who reported
that participants used attractiveness as a distinguishing cue between
real and AI-generated faces. These results highlight a possible
bidirectional and context-dependent influence of attractiveness on
reality judgments, of which the exact shape needs further investigation.

@@

The first sample includes the same participants as study 1 (N = 141, see
above). For the second sample (data and preprocessing information
available at https://github.com/RealityBending/FakeFace2), 248
participants were initially recruited through academic platforms (SONA
and SurveySwap). We removed 15 participants with data suggesting
low-effort responding (no variation in their responses indicative of
straightlining) as well as those (N=16) that did not believe in the
experimental manipulation and were fully confident that all images were
real or fake. Similarly to the first sample, we included all hetero and
bisexual participants and stimuli from the opposite gender, resulting in
a final sample included 189 participants (Mean age = 28.4, SD = 14.0,
range: {[}18, 69{]}; Sex: 76.2\% females). The study was approved by the
University of Sussex Ethics Board (ER/ST633/1).

@@

\begin{center}
\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{./figures/fig5.png}
\end{center}

@@

@@

\bibitem[\citeproctext]{ref-viola2023designed}
Viola, M., \& Voto, C. (2023). Designed to abuse? Deepfakes and the
non-consensual diffusion of intimate images. \emph{Synthese},
\emph{201}(1), 30.
